
```{r, setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  tidy = FALSE,
  error = FALSE,
  fig.width = 8,
  fig.height = 8)
```

# processx

> Execute and Control System Processes

[![Linux Build Status](https://travis-ci.org/MangoTheCat/processx.svg?branch=master)](https://travis-ci.org/MangoTheCat/processx)
[![Windows Build status](https://ci.appveyor.com/api/projects/status/github/MangoTheCat/processx?svg=true)](https://ci.appveyor.com/project/gaborcsardi/processx)
[![](http://www.r-pkg.org/badges/version/processx)](http://www.r-pkg.org/pkg/processx)
[![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/processx)](http://www.r-pkg.org/pkg/processx)

Portable tools to run system processes in the background,
read their standard output and error, kill and restart them.

---

  - [Features](#features)
  - [Installation](#installation)
  - [Usage](#usage)
    - [Starting processes](#starting-processes)
    - [Killing and restarting a process](#killing-and-restarting-a-process)
    - [Standard output and error](#standard-output-and-error)
    - [End of output](#end-of-output)
    - [Waiting on a process](#waiting-on-a-process)
	- [Exit statuses](#exit-statuses)
	- [Errors](#errors)
  - [License](#license)

## Features

* Start system processes in the background and find their
  process id.
* Check if a background process is running.
* Wait on a background process.
* Get the exit status of a background process, if it has already
  finished.
* Kill background processes, together with their children.
* Kill background process, when its associated object is garbage
  collected.
* Restart background processes.
* Read the standard output and error, using non-blocking connections.
* Portable, works on Linux, macOS and Windows.
* Lightweight, it only depends on the also lightweight
  [R6](http://r-pkg.org/pkg/R6) package.
* Contains no compiled code, so it is easy to install.

## Caveats

* There is no way of doing a *blocking* read from the standard
  output or error stream of the process.

## Installation

```{r eval = FALSE}
source("https://install-github.me/MangoTheCat/processx")
```

## Usage

Note: the following external commands are usually present in macOS and
Linux systems, but not necessarily on Windows.

```{r}
library(processx)
```

### Starting processes

`processx` provides two ways to start processes. The first one
requires a single command, and a character vector of arguments.
Both the command and the arguments will be shell quoted, so it is safe
to include spaces and other special characters:

```{r}
p1 <- process$new("sleep", "20")
```

The other way is to supply a full shell command line via the
`commandline` argument:

```{r}
p2 <- process$new(commandline = "sleep 20")
```

Both methods will run the specified command or command line via the
`pipe` base R function, which currently uses a shell to start them.

### Killing and restarting a process

A process can be killed via the `kill` method. This also kills
all child processes.

```{r}
p1$is_alive()
p2$is_alive()
p1$kill()
p2$kill()
p1$is_alive()
p2$is_alive()
```

A process can be restarted via `restart`. This works if the process
has been killed, if it has finished regularly, or even if it is running
currently. If it is running, then it will be killed first.

```{r}
p1$restart()
p1$is_alive()
```

### Standard output and error

By default the standard output and error of the processes are redirected
to temporary files. You can set the `stdout` and `stderr` constructor
arguments to `FALSE` to force discarding them, in case you have a long
running process with a lot of output. You can also set them to file names,
if you want the output and/or error streams in some specific files.

The `read_output_lines` and `read_error_lines` methods can be used
to read the standard output or error. They work the same way as the
`readLines` base function.

Note that the connections used for reading the output and error streams
are non-blocking text connections, so the read functions will return
immediately, even if there is no text to read from them.

```{r}
p <- process$new(commandline = "echo foo; >&2 echo bar; echo foobar")
p$read_output_lines()
p$read_error_lines()
```

To check if there is anything available for reading on the standard output
or error streams, use the `can_read_output` and `can_read_error` functions:

```{r}
p <- process$new(commandline = "echo foo; sleep 2; echo bar")

Sys.sleep(1)
## There must be output now
p$can_read_output()
p$read_output_lines()

## There is no more output now
p$can_read_output()
p$read_output_lines()

Sys.sleep(2)
## There is output again
p$can_read_output()
p$read_output_lines()

## There is no more output
p$can_read_output()
p$read_output_lines()
```

### End of output

There is no standard way in R to signal the end of a connection,
unfortunately. Most R I/O is blocking, and the end of file is reached when
nothing can be read from the connection. This clearly does not work for
non-blocking connections.

For `process` standard output and error streams, you can use the
`is_eof_output` and `is_eof_error` functions to check if there is any
chance that more output will arrive on them later.


### Waiting on a process

As seen before, `is_alive` checks if a process is running. The `wait`
method can be used to wait until it has finished. E.g. in the following
code `wait` needs to wait about 2 seconds for the `sleep` shell command
to finish.

```{r}
p <- process$new(commandline = "sleep 2")
p$is_alive()
Sys.time()
p$wait()
Sys.time()
```

It is safe to call `wait` multiple times:

```{r}
p$wait() # already finished!
```

### Exit statuses

After a process has finished, its exit status can be queried via the
`get_exit_status` method. If the process is still running, then this
method returns `NULL`.

```{r}
p <- process$new(commandline = "sleep 2")
p$get_exit_status()
p$wait()
p$get_exit_status()
```

Note that even if the process has finished, `get_exit_status` might report
`NULL` if `wait` was not called. If you know that the process has finished,
call `wait` first, and only then `get_exit_status`.

```{r}
p <- process$new(commandline = "true")

# wait until it surely finishes
Sys.sleep(1)
p$is_alive()
p$get_exit_status()

p$wait()
p$get_exit_status()
```

### Errors

Errors are typically signalled via non-zero exits statuses. `processx`
currently has no special behavior to handle errors.

If you want to make sure that the process has started successfully,
check that it is running via `is_alive()`, and if it is not, then check
its exit status (after calling `wait`).

```{r, error = TRUE}
p <- process$new("nonexistant-command-for-sure")
if (!p$is_alive()) {
  p$wait()
  if (p$get_exit_status() != 0) {
    stop("nonexistant-command-for-sure failed")
  }
}
```

## License

MIT © Gábor Csárdi
